---
title: "Kwara state density and vector data"
author: "Gerry Ryan"
format:
  html: default
  gfm: default
  pdf: default
editor: source
date: 2023-07-26
---


```{r, echo = FALSE, results = 'hide', message=FALSE}
library(terra)
library(viridisLite)
library(tidyterra)
library(ggplot2)
library(dplyr)
library(geodata)
library(RColorBrewer)

built_nigeria <- rast("output/grids/built_nigeria.tif")
built_kwara_3 <- rast("output/grids/urban_kwara_3.tif")
built_kwara_6 <- rast("output/grids/urban_kwara_6.tif")
an_ratios_kwara <- rast("output/grids/an_ratios_kwara.tif")
irs_ratios_kwara <- rast("output/grids/irs_ratios_kwara.tif")
itn_ratios_kwara <- rast("output/grids/itn_ratios_kwara.tif")
ir_2017_kwara <- rast("output/grids/ir_2017_kwara.tif")
irs_change_kwara <- rast("output/grids/irs_change_kwara.tif")
itn_change_kwara <- rast("output/grids/itn_change_kwara.tif")

ilorin_area <- ext(
  c(
    4.25,
    4.75,
    8.25,
    8.75
  )
)

nigeria_states <- gadm(
  country = "NGA",
  level = 1,
  path = "data/downloads"
)


kwara_vector <- nigeria_states %>%
  filter(NAME_1 == "Kwara")
```

## Urban density

### Built volume base layer
The basis for these  is the [GHS built-up volume grid](https://ghsl.jrc.ec.europa.eu/ghs_buV2023.php) produced by the European Commission Joint Research Council, which measures the bult volume in cubic metres over space.

#### Kwara state
```{r, echo = FALSE}
plot(
  built_nigeria %>%
    mask(kwara_vector) %>%
    crop(kwara_vector),
  col = mako(
    n = 2048,
   direction = -1
  )
)
lines(
  kwara_vector,
  col = "grey40"
)
```


#### Ilorin city
```{r, echo = FALSE}
plot(
  built_nigeria %>%
    mask(kwara_vector) %>%
    crop(kwara_vector),
  col = mako(
    n = 2048,
   direction = -1
  ),
  ext = ilorin_area
)
lines(
  kwara_vector,
  col = "grey40"
)
```

### Classification into density classes

#### Kwara state, 3 classes

```{r, echo = FALSE}
 plot(
  built_kwara_3,
  col = mako(
    n = 3,
    direction = -1
  )
)
lines(
  kwara_vector,
  col = "grey40"
)
```

#### Kwara state 6 classes
```{r, echo = FALSE}
plot(
  built_kwara_6,
  col = mako(
    n = 6,
    direction = -1
  )
)
lines(
  kwara_vector,
  col = "grey40"
)
```

#### Ilorin city 3 classes
```{r, echo = FALSE}
 plot(
  built_kwara_3,
  col = mako(
    n = 3,
    direction = -1
  ),
  ext = ilorin_area
)
lines(
  kwara_vector,
  col = "grey40"
)
```

#### Ilorin city six classes:
```{r, echo = FALSE}
plot(
  built_kwara_6,
  col = mako(
    n = 6,
    direction = -1
  ),
  ext = ilorin_area
)
lines(
  kwara_vector,
  col = "grey40"
)
```


## Proportional abundance of dominant vectors
These plots are from [*Modelling the relative abundance of the primary African vectors of malaria before and after the implementation of indoor, insecticide-based vector control*, Sinka et al. 2016](https://doi.org/10.1186/s12936-016-1187-8), cropped to Kwara state. The 

### Base proportional abundance of *Anopheles gambiae*, *An. funestus*, and *An gambiae*
This map shows the baseline relative proportional abundance of the three dominant vectors.
```{r, echo = FALSE}
plot(
  an_ratios_kwara,
  col = mako(
    n = 2048,
    direction = -1
  ),
  range = 0:1
)
```

### After IRS

#### Proportional relative abundance post-IRS
This shows the expected proportional abundance after the application of IRS
```{r, echo = FALSE}
plot(
  irs_ratios_kwara,
  col = mako(
    n = 2048,
    direction = -1
  ),
  range = 0:1
)
```

#### Change in proportion after IRS
This is the differnce between the base proportional relative abundance, and the post-IRS proportional abundance (the previous two maps). Where it is green, the species is expected to be relatively more dominant post-IRS, and where it is purple it is expected to be relatively less dominant post-IRS
```{r, echo = FALSE}
plot(
  irs_change_kwara,
  col = brewer.pal(
    n = 11,
    name = "PRGn"
  ),
  range = c(-1, 1)
)
```


### After ITN use

#### Proportional relative abundance post-ITN use
This shows the expected proportional abundance after the distribution of ITNs
```{r, echo = FALSE}
plot(
  itn_ratios_kwara,
  col = mako(
    n = 2048,
    direction = -1
  ),
  range = 0:1
)
```

#### Change in proportion after ITN use
This is the difference between the base proportional relative abundance, and the post-ITN proportional abundance (the first relative abundance map, and the previous map). Where it is green, the species is expected to be relatively more dominant post-IRS, and where it is purple it is expected to be relatively less dominant post-ITN distribution
```{r, echo = FALSE}
plot(
  irs_change_kwara,
  col = brewer.pal(
    n = 11,
    name = "PRGn"
  ),
  range = c(-1, 1)
)
```

## Insecticide resistance
These plots are from [*Mapping trends in insecticide resistance phenotypes in African malaria vectors*, Hancock et al 2020](https://doi.org/10.1186/s12936-016-1187-8), cropped to Kwara state.

### Alphacypermethrin
Predicted mean proportional mortality to **alphacypermethrin** in Kwara state. Low values (dark colours) indicate low mortality, meaning high resistance to this insecticide
```{r, echo = FALSE}
plot(
  ir_2017_kwara$Alphacypermethrin_mortality,
  col = mako(
    n = 2048
  ),
  range = 0:1
)
```

### DDT
Predicted mean proportional mortality to **DDT** in Kwara state. Low values (dark colours) indicate low mortality, meaning high resistance to this insecticide
```{r, echo = FALSE}
plot(
  ir_2017_kwara$DDT_mortality,
  col = mako(
    n = 2048
  ),
  range = 0:1
)
```

### Deltamethrin
Predicted mean proportional mortality to **deltamethrin** in Kwara state. Low values (dark colours) indicate low mortality, meaning high resistance to this insecticide
```{r, echo = FALSE}
plot(
  ir_2017_kwara$Deltamethrin_mortality,
  col = mako(
    n = 2048
  ),
  range = 0:1
)
```

### Lambdacyhalothrin
Predicted mean proportional mortality to **lambdacyhalothrin** in Kwara state. Low values (dark colours) indicate low mortality, meaning high resistance to this insecticide
```{r, echo = FALSE}
plot(
  ir_2017_kwara$Lambdacyhalothrin_mortality,
  col = mako(
    n = 2048
  ),
  range = 0:1
)
```

### Permethrin
Predicted mean proportional mortality to **permethrin** in Kwara state. Low values (dark colours) indicate low mortality, meaning high resistance to this insecticide
```{r, echo = FALSE}
plot(
  ir_2017_kwara$Alphacypermethrin_mortality,
  col = mako(
    n = 2048
  ),
  range = 0:1
)
```
