---
title: "Urban density classification for Nigeria"
author: "Gerry Ryan"
format:
  html: default
  gfm: default
  pdf: default
editor: source
date: 2023-07-26
---

This is a preliminary classification of urban density based on volume of built structures in an area for Nigeria.

```{r, echo = FALSE, results = 'hide', message=FALSE}
library(terra)
library(viridisLite)
library(tidyterra)
library(ggplot2)
library(dplyr)
library(geodata)

built_nigeria <- rast("output/grids/built_nigeria.tif")

built_nigeria_3 <- rast("output/grids/urban_nigeria_3.tif")
built_nigeria_6 <- rast("output/grids/urban_nigeria_6.tif")

built_kwara_3 <- rast("output/grids/urban_kwara_3.tif")
built_kwara_6 <- rast("output/grids/urban_kwara_6.tif")

lagos_area <-  ext(
  c(
    3,
    3.5,
    6.3,
    6.8
  )
)

ilorin_area <- ext(
  c(
    4.25,
    4.75,
    8.25,
    8.75
  )
)

nigeria_states <- gadm(
  country = "NGA",
  level = 1,
  path = "data/downloads"
)


kwara_vector <- nigeria_states %>%
  filter(NAME_1 == "Kwara")


lagos_ogun_vector <- nigeria_states %>%
  filter(NAME_1 == "Lagos" | NAME_1 == "Ogun")

ext_lagos_ogun <- ext(lagos_ogun_vector)


built_lagos_ogun_3 <- mask(
  x = built_nigeria_3,
  mask = lagos_ogun_vector
) %>%
  crop(
    y = lagos_ogun_vector
  )

built_lagos_ogun_6 <- mask(
  x = built_nigeria_6,
  mask = lagos_ogun_vector
) %>%
  crop(
    y = lagos_ogun_vector
  )

```

The basis for these classifications is the [GHS built-up volume grid](https://ghsl.jrc.ec.europa.eu/ghs_buV2023.php) produced by the European Commission Joint Research Council, which measures the bult volume in cubic metres over space.

## Base layer of built structure volume

Here is the base layer for all of Nigeria:

```{r, echo = FALSE}
plot(
  built_nigeria,
  col = mako(
    n = 2048,
   direction = -1
  )
)
lines(
  nigeria_states,
  col = "grey"
)

# ggplot() +
#   geom_spatraster(
#     data = built_nigeria
#   ) +
#   scale_fill_viridis_c(
#     option = "G",
#     direction = -1,
#     na.value = "white"
#   ) +
#   geom_spatvector(
#     data = nigeria_states,
#     colour = "grey40",
#     fill = NA
#   ) +
#   theme_minimal()
```

And focussed in on the south-west of the country in Lagos and Ogun states:

```{r, echo = FALSE}
plot(
  built_nigeria,
  col = mako(
    n = 2048,
   direction = -1
  ),
  ext = ext_lagos_ogun
)
lines(
  nigeria_states,
  col = "grey40"
)
```

Close up of Lagos city:

```{r, echo = FALSE}
plot(
  built_nigeria,
  col = mako(
    n = 2048,
   direction = -1
  ),
  ext = lagos_area
)
lines(
  nigeria_states,
  col = "grey40"
)
```

And here Kwara state only

```{r, echo = FALSE}
plot(
  built_nigeria %>%
    mask(kwara_vector) %>%
    crop(kwara_vector),
  col = mako(
    n = 2048,
   direction = -1
  )
)
lines(
  kwara_vector,
  col = "grey40"
)
```

And a close up of Ilorin city:

```{r, echo = FALSE}
plot(
  built_nigeria %>%
    mask(kwara_vector) %>%
    crop(kwara_vector),
  col = mako(
    n = 2048,
   direction = -1
  ),
  ext = ilorin_area
)
lines(
  kwara_vector,
  col = "grey40"
)
```

## Classification

Using Jenks natural breaks classification of the cubed root of volume^[This is done to 'flatten' the volumes to allowing for greater differentiation among the lower end of the scale, and relatively less differentiation at the upper end, e.g. among very tall buildings] we classify these into either three density classes:


```{r, echo = FALSE}
plot(
  built_nigeria_3,
  col = mako(
    n = 3,
    direction = -1
  )
)
lines(
  nigeria_states,
  col = "grey40"
)
```

Or six density classes:

```{r, echo = FALSE}
plot(
  built_nigeria_6,
  col = mako(
    n = 6,
    direction = -1
  )
)
lines(
  nigeria_states,
  col = "grey40"
)
```

### Lagos and Ogun states

This is the same data as above focussing on Lagos and Ogun states for 3 classes:

```{r, echo = FALSE}
 plot(
  built_nigeria_3,
  ext = ext_lagos_ogun,
  col = mako(
    n = 3,
    direction = -1
  )
)
lines(
  nigeria_states,
  col = "grey40"
)
```

And six classes:

```{r, echo = FALSE}
plot(
  built_nigeria_6,
  ext = ext_lagos_ogun,
  col = mako(
    n = 6,
    direction = -1
  )
)
lines(
  nigeria_states,
  col = "grey40"
)
```

### Lagos city

Close up of Lagos city for 3 classes:

```{r, echo = FALSE}
 plot(
  built_nigeria_3,
  ext = lagos_area,
  col = mako(
    n = 3,
    direction = -1
  )
)
lines(
  nigeria_states,
  col = "grey40"
)
```

And six classes:

```{r, echo = FALSE}
plot(
  built_nigeria_6,
  ext = lagos_area,
  col = mako(
    n = 6,
    direction = -1
  )
)
lines(
  nigeria_states,
  col = "grey40"
)
```

### Kwara state

Kwara state for 3 classes:

```{r, echo = FALSE}
 plot(
  built_kwara_3,
  col = mako(
    n = 3,
    direction = -1
  )
)
lines(
  kwara_vector,
  col = "grey40"
)
```

And six classes:

```{r, echo = FALSE}
plot(
  built_kwara_6,
  col = mako(
    n = 6,
    direction = -1
  )
)
lines(
  kwara_vector,
  col = "grey40"
)
```

### Ilorin city

And the area around Ilorin city state for 3 classes:

```{r, echo = FALSE}
 plot(
  built_kwara_3,
  col = mako(
    n = 3,
    direction = -1
  ),
  ext = ilorin_area
)
lines(
  kwara_vector,
  col = "grey40"
)
```

And six classes:

```{r, echo = FALSE}
plot(
  built_kwara_6,
  col = mako(
    n = 6,
    direction = -1
  ),
  ext = ilorin_area
)
lines(
  kwara_vector,
  col = "grey40"
)
```
